Analyzing DECREASE trials to determine evidence for data fabrication

### 2017-07-13 submission/manuscript.tex
Revision round
Incorporate revisions Jelte, Markus, and Gerben.

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Analyzing DECREASE trials for extent of data fabrication {[}working title{]}},
            pdfauthor={Chris HJ Hartgerink, Gerben ter Riet, Marleen Kemper},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}
  \title{Analyzing DECREASE trials for extent of data fabrication {[}working
title{]}}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{Chris HJ Hartgerink, Gerben ter Riet, Marleen Kemper}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{24 May, 2017}

@@

The reports on the integrity of the DECREASE trials primarily focused on
the provenance of the raw data but did not investigate the extent to
which the DECREASE trials deviated from comparable trials. The
provenance is primarily concerned with the origins of the data,
verifying things such as (but not limited to) the informed consent and
whether data corresponded to patient files. The committee reports did
not neglect statistical evaluation however: a statistical expert
evaluated the applicability of forensic statistical
methods\textsuperscript{6} to evaluate results of trials separately
(i.e., DECREASE-I, DECREASE-IV). Nonetheless, comparing across trials is
a method that has previously been used to monitor trial data quality or
to test for potential data anomalies\textsuperscript{11}. Moreover, this
method has previously proven to be effective in detecting data
fabrication\textsuperscript{12}. Comparing the DECREASE trials to other
published trials studying the effectiveness of beta-blockers with
respect to perioperative mortality could prove informative of the
potential extent of the fabrication in the DECREASE trials.

@@

The latest guidelines of the European Society of Cardiology and the
American Heart Association still recommend the usage of beta-blockers in
certain cases {[}@;@{]}. Therefore, we evaluate the usage of
beta-blockers with respect to its effectiveness for the the
perioperative period in patients undergoing non-cardiac surgery. First,
we estimated the extent of data fabrication in the DECREASE
studies\textsuperscript{2,3} Secondly, we evaluated the clinical studies
used in the different meta-analysis from a pharmacological point of
view.

\section{Part 1: statistical estimation of data fabrication in DECREASE
trials}\label{part-1-statistical-estimation-of-data-fabrication-in-decrease-trials}

To statistically investigate the evidence of data fabrication in the
DECREASE studies\textsuperscript{2,3}, we took three steps. First, we
reproduced the findings from the 2014 meta-analysis by Bouri et
al.\textsuperscript{9} that contained sufficient information to estimate
the deviation of the DECREASE trials from other published trials on
beta-blockers. We also include type of beta-blocker to inspect whether
this is predictive of the effect of beta-blockers on perioperative
mortality. Second, we evaluated the probability that the DECREASE trials
occurred assuming no data fabrication. Third, we reversed this
assumption and assumed that data fabrication did occur and estimated how
many data points would have to be fabricated to reproduce the results of
the DECREASE trials, if the other published trials are regarded as
estimating the true effect of beta-blockers on perioperative mortality
in patients undergoing non-cardiac surgery.

@@

\begin{verbatim}
## Warning in library(package, lib.loc = lib.loc, character.only = TRUE,
## logical.return = TRUE, : there is no package called 'metafor'
\end{verbatim}

\begin{verbatim}
## Installing package into '/home/chjh/R/x86_64-redhat-linux-gnu-library/3.3'
## (as 'lib' is unspecified)
\end{verbatim}

@@

Based on the non-DECREASE estimates from Step 1, we estimated the
probability of obtaining the results in the DECREASE trials. To this
end, we assumed that the non-DECREASE trials provide a valid
representation of the true effect of beta-blockers on perioperative
mortality (similar to Bouri et al.\textsuperscript{9}). The estimated
probability is also known as the veracity of the
data\textsuperscript{17}, which indicates the probability of the
observed data under a given true effect. We assumed that the
non-DECREASE studies estimated the true effect distribution of
perioperative beta-blockade on mortality, not perturbed by publication
bias due to statistical (non)significance. Publication bias was assumed
to not be a problem because a substantial number of nonsignificant
effects are included in the dataset (9 of 11 results are
nonsignificant).

@@

Based on the estimated effect distribution from the non-DECREASE trials,
we calculated the probability of each DECREASE trial result, or a more
extreme result. In other words, we computed the \emph{p}-value for the
null hypothesis that the DECREASE trials arise from the same effect
distribution as the non-DECREASE trials. This assumes that the
information available from the other trials is informative of the true
population effect.

@@

Results from Step 1 indicated that no between-trial variance (i.e.,
homogeneity; \(\tau^2=0\)) of the effects was observed; given the small
number of trials included (i.e., 9) this estimate is uncertain, however.
We conducted sensitivity analyses to see how dependent results are on
the heterogeneity estimate. The probability of observing the DECREASE
trials stays approximately below 1 out of 1000 until the variance
estimate is 0.25; the probability stays approximately below 1 out of 100
until the variance estimate is 0.43 (see Figure 2). To put these numbers
into context, a variance of 0.25 would suggest that results of
perioperative beta-blockade vary substantially even if perioperative
beta-blockade has no effect whatsoever (RRs between 0.779 and 1.284 in
64\% of the cases) all due to contextual circumstances of the study.

@@

For each hypothesis of \emph{X} out of \emph{N} fabricated data points,
we computed the probability that the fabricated data are more likely
than the genuine data (\(p_F=P(L_{fabricated}>L_{genuine})\)). Based on
\(p_F\), we computed the confidence interval for \(X\) (i.e.,
\(X_{LB};X_{UB}\)). For a 95\% confidence interval, the lowerbound is
equal to the \(p_F\) closest to .025, whereas the upperbound is equal to
the \(p_F\) closest to .975.

@@

Based on the meta-analytic effect from 2 and the cell sizes from 3, we
computed the likelihoods \(L_{fabricated}\) and \(L_{genuine}\) using
Equation \ref{eq1}. As mentioned before, we computed \(p_F\), which
indicated the probability that the data are more likely under the
estimates resulting from the (allegedly) fabricated data (i.e., the
DECREASE trials) than under the estimates resulting from the genuine
data (i.e., the non-DECREASE trials;
\(p_F=P(L_{fabricated}>L_{genuine})\)).

@@

\section{Part 2:}\label{part-2}

@@

The effect of beta-blockade on perioperative mortality was already
unclear following the findings of scientific misconduct; these results
strongly affirm that the DECREASE trials should be neglected when
assessing the effectiveness of beta-blockers. Our results indicate that
the results from the DECREASE trials are nearly impossible to have
arisen from the same effect inspected by the non-DECREASE trials, except
when we assume at least some of the data were manipulated. As such, the
scientific validity of the DECREASE-I and DECREASE-IV trials should by
now be clear: they are to be regarded as irrelevant when assessing the
effectiveness of beta-blockade on perioperative mortality. Nonetheless,
the original papers that presented these trial results are not yet
retracted\textsuperscript{2,3}, something we strongly recommend based on
the combination of our results and the integrity
reports\textsuperscript{4--6}.

@@

Nonetheless, references \emph{are} made without clear comments. Given
the confirmation of problems in the DECREASE-I and DECREASE-IV trials in
our results, it stresses that there is reason to distrust trials by
Poldermans. We pose that investigations should be initiated into works
where Poldermans was clearly involved and which were not cleared by the
scientific committees of Erasmus MC in their misconduct investigations.
Especially those papers cited as evidence in the ACC/AHA guidelines
should be investigated, considering they affect patients and their
treatment directly.

@@

The results of our analyses also highlight that, despite the lack of
availability of the raw data, summary results from larger samples allow
for better estimates of the number of fabricated data points when
similar trials are available. Moreover, larger trials result in
relatively more certainty (e.g., DECREASE-IV) about the estimated number
of fabricated data points, when using the inversion method, compared to
smaller trials (e.g., DECREASE-I). This increased certainty is due to
decreased standard errors at the population level of the estimated
effects, resulting in higher sensitivity to data anomalies. Nonetheless,
much residual uncertainty remains and simply fewer information is
available in the summary results. Raw data availability would improve
the options open to detect potential anomalies (note: raw data are
available for DECREASE VI, but upon a freedom of information request
Erasmus MC refused to share these data). The results also highlight that
in order to prevent detection, it would be in the fabricators' interest
to fabricate small studies even if raw data are hidden away (assuming
the fabricator wants to remain undetected).

@@

In sum, our research indicates that the DECREASE trials are nearly
impossible and contain at least some manipulated data points. We
recommend retraction of the DECREASE-I and DECREASE-IV trials, and
recommend renewed investigations into Poldermans' work --- especially
that work still referenced by guidelines on the use of beta-blockers.
Moreover, it remains unclear whether beta-blockers might be effective
given the right treatment. We recommend new and more extensively
controlled, confirmatory trials to determine whether there is any use in
administering beta-blockers in order to decrease perioperative mortality
--- at the moment there is insufficient evidence to determine any useful
effect of beta-blockers.

>>>>

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Analyzing DECREASE trials to determine evidence for data fabrication},
            pdfauthor={Chris HJ Hartgerink, Gerben ter Riet, Marleen Kemper},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}
  \title{Analyzing DECREASE trials to determine evidence for data fabrication}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{Chris HJ Hartgerink, Gerben ter Riet, Marleen Kemper}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{06 June, 2017}

@@

The reports on the integrity of the DECREASE trials primarily focused on
the provenance of the raw data but did not investigate the extent to
which the DECREASE trials deviated from comparable trials. Provenance is
primarily concerned with the origins of the data, verifying things such
as (but not limited to) the informed consent and whether data
corresponded to patient files. The committee reports did not neglect
statistical evaluation however: a statistical expert evaluated the
applicability of forensic statistical methods\textsuperscript{6} to
evaluate results of trials separately (i.e., DECREASE-I, DECREASE-IV).
Nonetheless, comparing across trials is a method that has previously
been used to monitor trial data quality or to test for potential data
anomalies\textsuperscript{11}. Moreover, this method has previously
proven to be effective in detecting data
fabrication\textsuperscript{12}. Comparing the DECREASE trials to other
published trials studying the effectiveness of beta-blockers with
respect to perioperative mortality could prove informative of the
potential extent of the fabrication in the DECREASE trials.

@@

To statistically investigate the evidence for data fabrication in the
DECREASE studies\textsuperscript{2,3}, we took three steps. First, we
reproduced the findings from the 2014 meta-analysis by Bouri et
al.\textsuperscript{9} that contained sufficient information to estimate
the deviation of the DECREASE trials from other published trials on
beta-blockers. We also include type of beta-blocker to inspect whether
this is predictive of the effect of beta-blockers on perioperative
mortality. Second, we evaluated the probability that the DECREASE trials
occurred assuming no data fabrication. Third, we reversed this
assumption and assumed that data fabrication did occur and estimated how
many data points would have to be fabricated to reproduce the results of
the DECREASE trials, if the other published trials are regarded as
estimating the true effect of beta-blockers on perioperative mortality
in patients undergoing non-cardiac surgery.

@@

@@

Based on the non-DECREASE estimates from Step 1, we estimated the
probability of obtaining the results in the DECREASE trials. To this
end, we assumed that the non-DECREASE trials provide a valid
representation of the true effect of beta-blockers on perioperative
mortality (similar to Bouri et al.\textsuperscript{9}). The estimated
probability of the observed data under a given true effect is also known
as the veracity of the data\textsuperscript{17}. We assumed that the
non-DECREASE studies estimated the true effect distribution of
perioperative beta-blockade on mortality, not perturbed by publication
bias due to statistical (non)significance. Publication bias was assumed
to not be a problem because a substantial number of nonsignificant
effects are included in the dataset (9 of 11 results are
nonsignificant).

@@

Based on the estimated effect distribution from the non-DECREASE trials,
we calculated the probability of each DECREASE trial result, or a more
extreme result. In other words, we computed the \(p\)-value for the null
hypothesis that the DECREASE trials arise from the same effect
distribution as the non-DECREASE trials. This assumes that the
information available from the other trials is informative of the true
population effect.

@@

Results from Step 1 indicated that no between-trial variance (i.e.,
homogeneity; \(\tau^2=0\)) of the effects was observed; given the small
number of trials included (i.e., 9) this estimate is uncertain, however.
We conducted sensitivity analyses to see how dependent results are on
the heterogeneity estimate (not preregistered). The probability of
observing the DECREASE trials stays approximately below 1 out of 1000
until the variance estimate is 0.25; the probability stays approximately
below 1 out of 100 until the variance estimate is 0.43 (see Figure 2).
To put these numbers into context, a variance of 0.25 would suggest that
results of perioperative beta-blockade vary substantially due to
contextual circumstances of the study, even if perioperative
beta-blockade has no effect whatsoever (RRs between 0.779 and 1.284 in
\textasciitilde{}64\% of the cases).

@@

For each hypothesis of \emph{X} out of \emph{N} fabricated data points,
we computed the probability that the fabricated data are more likely
than the genuine data (\(p_F=P(L_{fabricated}>L_{genuine})\)). Based on
\(p_F\), we computed the confidence interval for the estimated \(X\)
fabricated datapoints (i.e., \(X_{LB};X_{UB}\)). For a 95\% confidence
interval, the lowerbound is equal to the \(p_F\) closest to .025,
whereas the upperbound is equal to the \(p_F\) closest to .975.

@@

Based on the meta-analytic effect from 2 and the cell sizes from 3, we
computed the likelihoods \(L_{fabricated}\) and \(L_{genuine}\) using
Equation \ref{eq1}. As mentioned before, we computed \(p_F\), which
indicates the probability that the data are more likely under the
estimates resulting from the (allegedly) fabricated data (i.e., the
DECREASE trials) than under the estimates resulting from the genuine
data (i.e., the non-DECREASE trials;
\(p_F=P(L_{fabricated}>L_{genuine})\)).

@@

@@

The effect of beta-blockade on perioperative mortality was already
unclear following the findings of scientific misconduct; these results
strongly affirm that the empirical results from the DECREASE trials
should be neglected when assessing the effectiveness of beta-blockers.
Our results indicate that the results from the DECREASE trials are
nearly impossible to have arisen from the same effect inspected by the
non-DECREASE trials, except when we assume at least some of the data
were manipulated. As such, the scientific validity of the DECREASE-I and
DECREASE-IV trials should by now be clear: they are to be regarded as
irrelevant when assessing the effectiveness of beta-blockade on
perioperative mortality. Nonetheless, the original papers that presented
these trial results are not yet retracted\textsuperscript{2,3}, despite
the integrity reports\textsuperscript{4--6}.

@@

Nonetheless, references are made without clear comments. Given the
confirmation of problems in the DECREASE-I and DECREASE-IV trials in our
results, it stresses that there is reason to distrust trials by
Poldermans. For the integrity of the guidelines and the safety of the
patients, we pose that investigations should be initiated into works
where Poldermans was clearly involved and which were not cleared by the
scientific committees of Erasmus MC in their misconduct investigations.
Especially those papers cited as evidence in the ACC/AHA guidelines
should be investigated, considering they affect patients and their
treatment directly.

@@

The results of our analyses also highlight that, despite the lack of
availability of the raw data, summary results from larger samples allow
for better estimates of the number of fabricated data points when
similar trials are available. Moreover, larger trials result in
relatively more certainty (e.g., DECREASE-IV) about the estimated number
of fabricated data points, when using the inversion method, compared to
smaller trials (e.g., DECREASE-I). This increased certainty is due to
decreased standard errors at the population level of the estimated
effects, resulting in higher sensitivity to data anomalies. Nonetheless,
much residual uncertainty remains and simply fewer information is
available in the summary results. Raw data availability would improve
the options open to detect potential anomalies (note: raw data are
available for DECREASE VI, but upon a freedom of information request by
the first author, Erasmus MC refused to share these data). The results
also highlight that in order to prevent detection, it would be in the
fabricators' interest to fabricate small and unreliable studies even if
raw data are hidden away (assuming the fabricator wants to remain
undetected), which ultimately detracts from the value of such a study
and hence the reward for fabrication.

@@

Moreover, the DECREASE and non-DECREASE trials seem to apply
beta-blockade from different conceptual viewpoints. The non-DECREASE
trials seem to focus purely on the application of beta-blockers in
itself, whereas the DECREASE trials use beta-blockade as a proxy to
decrease resting heart rate\textsuperscript{2,3}. As such, the DECREASE
studies applied beta-blockade at least a week in advance, specifically
in order to lower patient's resting heart rate to \textless{}70BPM and
potentially habituate the patient to the effects of the beta-blockade.
Other studies apply the beta-blockade just prior to the surgery
(maximum: one day prior) and therefore seem to regard the treatment
specifically and not the proxy of lowered BPM. As such, the differences
between the DECREASE and non-DECREASE trials might also in part be a
consequence of the different approaches in the various trials. Whether
these differences matter in treatment decisions is worthy of further
research in a clinical trial with high statistical power to find such
differences.

In sum, our research indicates that the DECREASE trials are nearly
impossible if we assume they investigate exactly the same effect as the
non-DECREASE trials and under that assumption, our results provide some
evidence that at least some data points were manipulated. However, these
differences might also be due to different conceptual approaches to how
beta-blockade might prevent mortality in non-cardiac surgery. We
recommend renewed investigations into Poldermans' work given these
findings --- especially those works still referenced by guidelines on
the use of beta-blockers without proper notice. Moreover, it remains
unclear whether beta-blockers might be effective in preventing mortality
rates in non-cardiac surgery patients. Considering this, we recommend
new and more extensively controlled, confirmatory trials to determine
whether there is any use in administering beta-blockers in order to
decrease perioperative mortality --- at the moment there is insufficient
evidence to determine any positive effect of beta-blockers on mortality
rates.
